# Introduction

In this vignette, we will explore the functions in the LatticeKrig package and show examples of how they can be used to solve problems. The LatticeKrig (LK) model is an example of the spatial statistics method known as kriging, adapted to large data sets.

## What is kriging?

  By spatial data, we mean the data contains the observed variable and
its location, the variable depends on the location, and  observations that are closer  are  more
similar values than those farther apart.

Two  quotes by Waldo Tobler sum this up nicely:
\begin{quote}
\textit{Everything is related to everything else,
           but near things are more related than distant things} 
  
\textit{The phenomenon external to an area of interest affects what goes on inside}.
\end{quote}  
In the methods of this package, we quantify this relatedness  of "near things"",  statistically using correlation.

Kriging (named for South African statistician Danie Krige) is a method for making predictions from a
spatial data set first developed for mineral exploration.
As such, kriging can be applied to a variety of important
data sets, from geological data to atmospheric data. Moreover, the method 
generalizes to the coordinates of a surface that need not be geographic or 2D/3D physical locations.
The classic method uses a model for the covariance (correlation) of the observations as a function 
of their distance of separation.
The Kriging predictions are weighted sums of the observations, where the weights depend on the
covariances and are different at every location. 

The standard spatial model for Kriging relates the observation to a sum of three components:
a \textit{fixed part} typcially a polynomial function of the locations (and covariates,
if provided),
a \textit{spatial process}, and \textit{measurement error}. This is a more practical model for data
analysis but involves more complicated formulae for the Kriging weights. The LatticeKrig (LK) model
like most spatial methods, decomposes the data into these three components.

## The LatticeKrig model

The key feature of LatticeKrig is that we model a spatial process as the sum of radial basis functions (functions that are symmetric around their center and are 0 for far away points) scaled by coefficients, which we assume are correlated. The smooth basis functions and correlated coefficients create a smooth function representation for the spatial process, and the structure of the basis functions and covariance has some flexibility so you can change the structure to make a more reasonable model for a certain problem. The linear polynomial in the locations and covariates is determined using generalized least squares, following the standard approach of Universal Kriging. To approximate the spatial process, we then fit the basis functions to the residuals from the linear model. In terms of linear algebra, the model is
\[\mathbf y = X \mathbf d + \Phi \mathbf c + \mathbf e\]
where $\mathbf y$ is the vector of variable measurements, $X$ is the matrix of locations and covariates, $\mathbf d$ is the vector of coefficients for the linear model, $\Phi$ is the matrix of basis functions evaluated at the data points, $\mathbf c$ is the vector of coefficients for each basis function, and $\mathbf e$ is the measurement error. If we let $\mathbf g$ represent the true values of the variable (without measurement error $\mathbf e$), we can unroll the matrix multiplications into sums and get the form
\[ g(\mathbf s) = \sum_{k=1}^n \phi_k(\mathbf s) \hat {\mathbf d}_k + \sum_{k=1}^m \psi_k (\mathbf s) \hat{\mathbf c}_k\]
We show the derivations of the equations for $\mathbf c$ and $\mathbf d$ in Appendix A, and show how all of these calculations are done in Appendix C.

The package is named LatticeKrig because of the placement of the basis functions: they are equally spaced in all dimensions on a lattice. We can also consider multiple different lattice sizes simultaneously to better capture different levels of resolution; by default, each additional level has half as much space between the basis functions in each dimension.
The following plot shows the basis functions for each of the three default levels.

But before we start with code here are the R, fields package, and LatticeKrig package versions
used throughout this vignette. Use these to check your versions if you get errors and we will try to make this vignette compatible with latter versions.   

```{r versions, echo=TRUE}
R.Version()$version.string
packageVersion( "fields")
packageVersion( "LatticeKrig")
```

```{r levelsPlot, echo=FALSE}

phi <- function(d) {
  return(1/3 * (1-d)^6 * (35*d^2 + 18*d + 3) * (d < 1))
}
overlap <- 2.5
gridPoints <- seq(0, 10, length=1000)
centers1 <- seq(0,10, length=11)
centers2 <- seq(0,10, length=21)
centers3 <- seq(0,10, length=41)
dists1 <- rdist(gridPoints, centers1)
dists2 <- rdist(gridPoints, centers2)
dists3 <- rdist(gridPoints, centers3)
values1 <- phi(dists1 / overlap)
values2 <- 1/2 * phi(2 * dists2 / overlap)
values3 <- 1/4 * phi(4 * dists3 / overlap)
matplot(x = gridPoints, cbind(values1, values2, values3), type="l", xlab="Location",
        ylab = "Basis function value", main="Basis Functions for 3 Levels", lty = 1,
        col = c(rep("black", 11), rep("red", 21), rep("green", 41)))
legend("topright", c("First Level", "Second Level", "Third Level"),
       col=c("black", "red", "green"), lty=1, bg="white")
#lines(values[,7], x=gridPoints, type="l", col="black", lwd=3)
```

To represent a curve, we multiply each basis function by a coefficient and add them together, as shown in the following plot.The blue bars show the coefficient value at each lattice point, and the black line shows the resulting curve from scaling the basis functions with these coefficients and adding them together.

```{r IntroCurveSimulations, fig.height=3.5, fig.width=6, echo=FALSE}
info <- LKrigSetup(as.matrix(c(0,10)), alpha = 1, nlevel = 1, NC = 11, NC.buffer = 0, a.wght=2.5, LKGeometry = "LKInterval")
set.seed(483765)
coefs <- LKrig.sim(seq(0,10,,100), info, M=1, just.coefficients = TRUE)
curve <- values1 %*% coefs
plot(curve, type="l", x = seq(0,10,,1000), xlab = "Location",
     ylab = "Curve value", main="Example 1-D Curve and Coefficients")
abline(v = 0:10, col="gray", lty=2)
abline(h = 0, col = "darkgray")
barplot(coefs, col = "lightblue", width = 0.2, space = c(-0.5, rep(4, 10)), add = TRUE)
```

## Glossary of important package functions

* \ttt{LatticeKrig}: A top level function that sets up the default spatial model, estimates some key spatial parameters, and uses the \ttt{LKrig} function for the kriging computation. \ttt{LatticeKrig} can use a minimal set of inputs and is a quick way to fit a kriging model to data.
* \ttt{LKrig}: Performs the Kriging computation and evaluates the Gaussian spatial likelihood function for a fixed LatticeKrig model. This is the main computational step in the package, and is common to all choices of geometries and models.
* \ttt{LKrigSetup}: Creates an \ttt{LKinfo} object, which is used to describe the complete spatial model for a \ttt{LatticeKrig} or \ttt{LKrig} call; especially useful for examining the effect of changing one parameter on the model.
* \ttt{surface}: Plots a fitted surface in 2D space as a color plot and adds contour lines.
* \ttt{image.plot}: Plots a dataset or fitted surface in 2D space as an image plot and adds a color bar legend.
* \ttt{predictSurface}: Takes a Kriging model and evaluates its fitted function on a grid of locations.
