# Quick Start Guide

In this section, we will lay out the bare essentials of the package as a quick overview. To fit a surface and interpolate data using \ttt{LatticeKrig}, the only required arguments are, naturally, the measurement locations (formatted in a matrix where each row is one location) and measurement values. However, we highly recommend using some of the optional parameters to customize the model to your specific data problem - several ways to do this are illustrated in this vignette. Calling the \ttt{LatticeKrig} function and passing in the locations and values will produce an \ttt{LKrig} object that contains all the information needed to predict the variable at any location. 

For a simple, 1-dimensional example, we will take our locations to be 50 randomly spaced points on the interval $[-6, 6]$, and our variable measurements to be the values of $\sin(x)$ at these locations. The goal of our kriging fit is to estimate this smooth curve from the observations.

## Fitting the model in one dimension

```{r QuickStart1D}
set.seed(223)
locations <- runif(50, min=-6, max=6)
locations <- as.matrix(locations)
observations <- sin(locations) + rnorm(50, sd = 1e-1)
kFit1D <- LatticeKrig(locations, observations)
```

Now we will print out the \ttt{LKrig} object: this list features the data's estimated covariance scale \ttt{rho} and estimated standard measurement error \ttt{sigma}, and many basis function settings: the type of basis function, how distance is measured, and the number and spacing of basis functions. In this example, all of this information is determined by \ttt{LatticeKrig} from the defaults, but can be changed with optional parameters.

```{r QuickStart1DPrintout}
print(kFit1D)
```

##Plotting the results

Now, we'll make a plot of the original 50 data points and the true function ($\sin(x)$) and the \ttt{LatticeKrig} fit at 200 equally spaced points to compare them.
```{r QuickStart1DPlot}
xGrid <- seq(-2*pi, 2*pi, len=200)
prediction <- predict(kFit1D, xGrid)
plot(locations, observations, main="1-Dimensional LatticeKrig Example", 
     xlab="Location", ylab="Measured Value")
lines(xGrid, sin(xGrid), col='blue')
lines(xGrid, prediction, col='red', lty=2, lwd=2)
```

For this example, the fitted curve (in red) matches the true function (in blue) rather closely. 

## Inference and Error Analysis

This next plot adds a collection of simulations based on the fitted curve and the 95\% confidence envelope for the fitted curve; the simulations get farther apart, meaning the confidence envelope gets wider, where there aren't many data points and especially at the edges of the region. These simulations are easier to compute than the standard error, so they are used to estimate the standard error of a model in LatticeKrig.

```{r QuickStart1DResidualPlot}
uncertainty <- LKrig.sim.conditional(kFit1D, x.grid = as.matrix(xGrid), M=50)
sdVec <- apply(uncertainty$g.draw, 1, sd)
norm <- (uncertainty$g.draw - as.vector(prediction)) / sdVec
gammas <- apply(abs(norm), 2, max)
gamma <- quantile(gammas, 0.95)
upperBound <- prediction + gamma*sdVec
lowerBound <- prediction - gamma*sdVec
plot(xGrid, upperBound, type="l", ylim = c(min(lowerBound), max(upperBound)),
     main = "Simulated Curves with Confidence Envelope", xlab = "Location", ylab = "Measured Value")
EnvelopePlot(xGrid, lowerBound, xGrid, upperBound)
for(i in 1:10) {
  lines(uncertainty$x.grid, uncertainty$g.draw[,i], col="gray")
}

```

## Fitting the model in two dimensions

For another, more practical example, we will predict the average spring temperature for locations throughout Colorado. Using the data set \ttt{COmonthlyMet}, we can make a surface showing our predictions over a range of longitudes and latitudes, use the \ttt{US} function to draw in the USA state borders to show where Colorado is, and draw the points where data was recorded. Notice that \ttt{LatticeKrig} will automatically discard any data points with missing values (NAs) if needed.

```{r QuickStart2D}
data(COmonthlyMet)
locations <- CO.loc
observations <- CO.tmean.MAM.climate
kFitWeather <- LatticeKrig(locations, observations)
surface(kFitWeather, main = "2-Dimensional LatticeKrig Example",
        xlab="Longitude", ylab="Latitude")
points(locations, pch = '*')
US(add=TRUE, col='black', lwd=4)
```

This plot is useful, but we can do better. We can see that the coldest temperatures are in the Rocky Mountains, which is not surprising. Thus, we might expect that we will get a more accurate fit by having \ttt{LatticeKrig} account for the elevation at each location as well. Another way we can improve the plot is by increasing its resolution - the current image is somewhat pixelated. The \ttt{surface} function will evaluate the surface at more points if we increase the \ttt{nx} and \ttt{ny} arguments: setting \ttt{nx=200, ny=150} will produce a grid of 30,000 points, which will take longer to compute but produces a nicer looking, more detailed plot. Finally, we can also have \ttt{surface} extend the evaluation all the way to the corners of the window by using the \ttt{extrap} argument; by default it doesn't extrapolate outside of the existing data, since the error often increases dramatically when predicting outside of the given data. However, extending the plot to the corners will make it look nicer. For the sake of example, we will also change the color scale in the image by setting the \ttt{col} parameter.

```{r QuickStart2DImproved}
data(COmonthlyMet)
locations <- CO.loc
observations <- CO.tmean.MAM.climate
elevations <- CO.elev
kFitWeather <- LatticeKrig(locations, observations, Z=cbind(elevations))
# look at the help file in fields for information on the grid.list format
prediction <- predictSurface(kFitWeather, grid.list = CO.Grid, ZGrid = CO.elevGrid,
                             nx = 200, ny = 150, extrap = TRUE)
surface(prediction, main = "Improved 2-Dimensional LatticeKrig Example", 
        xlab="Longitude", ylab="Latitude", col=larry.colors())
US(add=TRUE, col='black', lwd=4)
```

This surface is so rough because it accounts for elevation; we can see that the plot is fairly smooth in the eastern half of the state, and extremely rough in the mountains. 

Finally, it is important to note some potential issues that \ttt{LatticeKrig} calculations won't catch. Because \ttt{LatticeKrig} estimates some parameters of the data, the model could be a poor fit if the estimates aren't reasonable. See section 6.4 for more details on this. The \ttt{LatticeKrig} model also approximates a thin plate spline by default, which may not be a good fit for a given problem. Finally, as with other curve fitting techniques, you should examine the residuals of the model for any patterns or features that may indicate a poor fit.

## Simulating a process from the LatticeKrig model
 
As a final topic we describe how to generate realizations from the Gaussian model in this package. The \ttt{LKinfo} object has a full description of the model and so simulation is easy. In the two examples of this section this object was set up from the top level function \ttt{LatticeKrig} and is the LKinfo component of the returned results. For more control over the model, however,  we recommend that this object be created separately. (See Section 3)

For 
\ttt{kFit1D} from Section 2.1 note that a listing of the full model is shown from.
```{r }
print(kFit1D$LKinfo)
```

Here we simulate 4 sample curves from this model and evaluate them on a finer grid of points than the observations. The random seed is set to reproduce these particular psuedo random draws. 

```{r }
set.seed(123)
gSim <- LKrig.sim(xGrid, kFit1D$LKinfo, M=4)
matplot(xGrid, gSim, type="l", lty=1, xlab="x", ylab="g(x)")
title("Simulated using model in 1D example")
```
Note that in actually fitting the data a linear function is also included but since this is not a random component it is not part of the simulated process. Also the variance of the process is set to one. 

These simulated curves are referred to as *unconditional* because they are
unrelated to the actual data except in terms of the range of the x values.
Another form of simulation is to generate the process conditional on
the observed data. This technique turns out to be very useful for quantifying the uncertianty in the curve estimate. The example below creates 25 draws from fitting the 1D example and to highlight the variability in these draws the plot is restricted to just a subset of the range of the data. This function returns several different parts of the estimate and so a list format is used. Note the use of the predict function to recover the estimated curve and also that the data is part of the fitted object. Within the range of the data all the conditional curves tend to track the estimate and the data, however, as one might expect beyond on the range of the observations there is much more variability among the simulated curves.   

```{r }
set.seed(123)
gCondSim <- LKrig.sim.conditional(kFit1D, M=25, x.grid=as.matrix(xGrid))
matplot(xGrid, gCondSim$g.draw, type="l", lty=1,
        xlab="x", ylab="y", col="grey", lwd=.5)
lines(xGrid, predict(kFit1D, xGrid), col="red")
points(kFit1D$x, kFit1D$y, pch=16, col="black")
```

## Extra Credit 

Here is a final example illustrating the power of deteriming the unceratinty by Monte Carlo. We generate a larger conditional sample over the subinterval [-3,0], find the minimum of each realization and plot the minimum and its location. This two dimensional distribution of minima and their locations is a valid approach to approximate the uncertainty of the estimated minimum of the true curve in this range. Moreover, it would be difficult to derive an analytic formula for this distribution. 

```{r }
xGrid2 <- as.matrix(seq(-3, 0, length.out=100))
set.seed(333)
suppressMessages(
  gCondSim <- LKrig.sim.conditional(kFit1D, x.grid=xGrid2, M=75)
)
# index of where minimum occurs in each draw
minXIndex <- apply(gCondSim$g.draw, 2, which.min)
XMin <- xGrid2[minXIndex]
YMin <- apply(gCondSim$g.draw, 2, min)
matplot(xGrid2, gCondSim$g.draw, type="l", lty=1,
        xlab="x", ylab="y", col="grey", ylim=c()) 
points(XMin, YMin, col="red4", cex=.5)         
lines(xGrid2, predict(kFit1D, xGrid2), col="red")
points(kFit1D$x,kFit1D$y, pch=16, col="black")        
```