# Common Error Messages and Frequently Asked Questions

## Could not find function

Make sure that the library is installed (\ttt{install.packages("LatticeKrig")}) and loaded (\ttt{library(LatticeKrig)}).

## Need to specify NC for grid size

## Invalid 'times' argument

## Only one alpha specifed for multiple levels

## Missing value where TRUE/FALSE needed

All of these errors can be caused by using \ttt{LKrig} instead of \ttt{LatticeKrig}. The \ttt{LatticeKrig} function has ways to either supply defaults or estimate all of the optional parameters that \ttt{LKrig} doesn't, so \ttt{LKrig} will produce errors like the ones above while \ttt{LatticeKrig} will work correctly. 

## Non-conformable arguments

This error can occur when using \ttt{LKrigSetup} (or \ttt{LatticeKrig}, by extension) on a 1-dimensional problem if you don't explicitly set \ttt{LKGeometry = "LKInterval"} for \ttt{LKrigSetup}. More generally, this problem occurs when trying to multiply matrices with incompatible dimensions.

## Argument is of length zero

This error most commonly occurs when using \ttt{LKrigSetup} in one dimension and passing in the range of the data explicitly. For example, \ttt{LKrigSetup(c(0,1), ...)} will cause this error (assuming the other arguments are provided correctly). The issue is that the \ttt{c} function doesn't format the input as a matrix, which is the format \ttt{LKrigSetup} expects. To fix this, just call \ttt{as.matrix} on the first parameter you give to \ttt{LKrigSetup} - so we would correct the example above to \ttt{LKrigSetup(as.matrix(c(0,1)), ...)}.

## Does the order of the parameters matter?

The order of the parameters only matters when you pass them in without specifying their names: for example \ttt{LatticeKrig(locations, values)} works, but \ttt{LatticeKrig(values, locations)} doesn't. However, if the names are specified, either order will work correctly: both \ttt{LatticeKrig(x = locations, y = values)} and \ttt{LatticeKrig(y = values, x = locations)} work as intended. The optional parameters can also be listed without their names, but then they would need to be in the correct order and every single one would need to be specified, so it is highly recommended to include the names alongside each optional parameter. For this reason, it is best practice to use the names of the parameters while passing them in, except in cases where it is obvious.

## The predicted values from my Kriging fit are nowhere near the data; what's wrong?

If your model includes covariates (the \ttt{Z} parameter of \ttt{LatticeKrig} and \ttt{LKrig}), your plot may not have included the effect of the covariate. The following code demonstrates this issue using the Colorado temperature data and kriging fit from the quick start guide, and how to fix the issue. Using the \ttt{surface} function will leave out the covariate, resulting in a plot that doesn't match the original data and is smoother than we might expect.

```{r FAQMissingCovariateBadPlot}
surface(kFitWeather, nx = 200, ny = 150, extrap = TRUE, main="Plot missing covariate",
        xlab = "Longitude", ylab="Latitude")
US(add=TRUE, col='black', lwd=4)
```

To fix this, call \ttt{surface} on a \ttt{predictSurface} object instead of on an \ttt{LKrig} object, and make sure to pass in the \ttt{grid.list} and \ttt{ZGrid} parameters  to the \ttt{predictSurface} call.

```{r FAQMissingCovariateGoodPlot}
prediction <- predictSurface(kFitWeather, grid.list = CO.Grid, 
                  ZGrid = CO.elevGrid, nx = 200, ny = 150, extrap = TRUE)
surface(prediction, main="Plot with covariate")
US(add=TRUE, col='black', lwd=4)
```

## Why aren't the settings in my LKrigSetup object being used by the kriging fit?

First, make sure everything is spelled correctly; R variables are case sensitive. For example, \ttt{LatticeKrig(x, y, LKInfo = info)} will not work, because the 'i' in "LKinfo" must be lowercase. Next, make sure that every parameter is being set correctly: in particular, don't confuse \ttt{x} with \ttt{X} or \ttt{alpha} with \ttt{a.wghts}. Also make sure that parameters that need to be passed as strings are in quotes, e.g. \ttt{LKGeometry = "LKSphere", distance.type="GreatCircle"}. If everything is set correctly and spelled correctly, make sure that the list from LKrigSetup is being passed in to your \ttt{LatticeKrig} or \ttt{LKrig} call.


<!--
aVec <- c(1, 0.25, 0.25*0.25)
 info <- LKrigSetup(newLocations, nlevel = 3, alpha = aVec/sum(aVec), LKGeometry = "LKSphere")
 kFit <- LatticeKrig(newLocations, dataValues, LKinfo=info)
Error in solve.default(t(X) %*% X, c(1, 0, 0)) : 
  Lapack routine dgesv: system is exactly singular: U[2,2] = 0
Timing stopped at: 0.05 0 0.06
Timing stopped at: 0.28 0 0.29 
-->

<!--
> info <- LKrigSetup(newLocations, nlevel = 3, alpha = aVec/sum(aVec), LKGeometry = "LKSphere")
> info <- LKrigSetup(dataLocations, nlevel = 3, alpha = aVec/sum(aVec), LKGeometry = "LKSphere")
Error in mLevel[l] <- nrow(grid.all.levels[[l]]) : 
  replacement has length zero
-->