# Frequently Asked Questions

## Does the order the parameters are listed in matter?

The order of the parameters only matters when you pass them in without specifying their names: for example \ttt{LatticeKrig(locations, values)} works, but \ttt{LatticeKrig(values, locations)} doesn't. However, if the names are specified, either order will work correctly: both \ttt{LatticeKrig(x = locations, y = values)} and \ttt{LatticeKrig(y = values, x = locations)} work as intended. The optional parameters can also be listed without their names, but then they would need to be in the correct order and every single one would need to be specified, so it is highly recommended to include the names alongside each optional parameter. 

## The predicted values from my Kriging fit are nowhere near the data; what's wrong?

If your model includes covariates (the \ttt{Z} parameter of \ttt{LatticeKrig} and \ttt{LKrig}), your plot may not have included the effect of the covariate. The following code demonstrates this issue using the Colorado temperature data and how to fix it; first, we will set up the model.
```{r FAQMissingCovariateSetup}
data(COmonthlyMet)
locations <- CO.loc
observations <- CO.tmean.MAM.climate
elevations <- CO.elev
kFit <- LatticeKrig(locations, observations, Z=cbind(elevations))
```

Using the \ttt{surface} function will leave out the covariate, resulting in a plot that doesn't match the original data and is smoother than we might expect.

```{r FAQMissingCovariateBadPlot}
surface(kFit, nx = 200, ny = 150, extrap = TRUE, main="Plot missing covariate",
        xlab = "Longitude", ylab="Latitude")
US(add=TRUE, col='black', lwd=4)
```

To fix this, call \ttt{surface} on a \ttt{predictSurface} object instead of on an \ttt{LKrig} object, and make sure to pass in the \ttt{grid.list} and \ttt{ZGrid} parameters  to the \ttt{predictSurface} call.

```{r FAQMissingCovariateGoodPlot}
prediction <- predictSurface(kFit, grid.list = CO.Grid, 
                  ZGrid = CO.elevGrid, nx = 200, ny = 150, extrap = TRUE)
surface(prediction, main="Plot with covariate")
US(add=TRUE, col='black', lwd=4)
```

## Why aren't the settings in my LKrigSetup object aren't being used by the kriging fit?

First, make sure everything is spelled correctly; R variables are case sensitive. For example, \ttt{LatticeKrig(x, y, LKInfo = info)} will not work, because the 'i' in "LKinfo" must be lowercase. Next, make sure that every parameter is being set correctly: in particular, don't confuse \ttt{x} with \ttt{X} or \ttt{alpha} with \ttt{a.wghts}. Also make sure that parameters that need to be passed as strings are in quotes, e.g. \ttt{LKGeometry = "LKSphere", distance.type="GreatCircle"}. If everything is set correctly and spelled correctly, make sure that the list from LKrigSetup is being passed in to your \ttt{LatticeKrig} or \ttt{LKrig} call.


<!--
aVec <- c(1, 0.25, 0.25*0.25)
 info <- LKrigSetup(newLocations, nlevel = 3, alpha = aVec/sum(aVec), LKGeometry = "LKSphere")
 kFit <- LatticeKrig(newLocations, dataValues, LKinfo=info)
Error in solve.default(t(X) %*% X, c(1, 0, 0)) : 
  Lapack routine dgesv: system is exactly singular: U[2,2] = 0
Timing stopped at: 0.05 0 0.06
Timing stopped at: 0.28 0 0.29 
-->

<!--
> info <- LKrigSetup(newLocations, nlevel = 3, alpha = aVec/sum(aVec), LKGeometry = "LKSphere")
> info <- LKrigSetup(dataLocations, nlevel = 3, alpha = aVec/sum(aVec), LKGeometry = "LKSphere")
Error in mLevel[l] <- nrow(grid.all.levels[[l]]) : 
  replacement has length zero
-->